Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-19 ms/ms.qmd
cleaning up

<!-- Paragraph 4+, hypotheses and predictions -->
The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers. We quantify this benefit as the amphistomy advantage (\aax{}), the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf [@parkhurst_adaptive_1978; @triplett_amphistomy_2024]. Why would \aax{} be greater in sun than shade? We consider three nonmutually exclusive hypotheses that we classify as 'acclimation', 'plasticity', and 'constitutive' ([Fig. @fig-concept]).

@@

```{r, concept, echo = FALSE}
#| label: fig-concept
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

knitr::include_graphics("../figures/amphisomy draft 2024-03-28.jpg")
```

>>>>

<!-- Paragraph 4+, hypotheses and predictions -->
The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers. We quantify this benefit as the amphistomy advantage (\aax{}), the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf [@parkhurst_adaptive_1978; @triplett_amphistomy_2024]. Why would \aax{} be greater in sun than shade? We consider three nonmutually exclusive hypotheses that we classify as 'acclimation', 'plasticity', and 'constitutive' ([Fig. @fig-hypotheses]).

@@

```{r, hypotheses, echo = FALSE}
#| label: fig-hypotheses
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

knitr::include_graphics("../figures/hypotheses.pdf")
```

### 2025-08-19 ms/ms.tex
cleaning up

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers. We quantify this benefit as
the amphistomy advantage (\aax{}), the log-response ratio of
photosynthesis in an amphistomatous leaf compared to an otherwise
identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
\aax{} be greater in sun than shade? We consider three nonmutually
exclusive hypotheses that we classify as `acclimation', `plasticity',
and `constitutive' (Fig.~\ref{fig-concept}).

@@

\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}

@@

\caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimatopn hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdqty{2000}) than low measurement light
intensity (\ppfdequals{150}), regardless of growth light intensity or
native plant area index (PAI). The plasticity hypothesis predicts that
\aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}

>>>>

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers. We quantify this benefit as
the amphistomy advantage (\aax{}), the log-response ratio of
photosynthesis in an amphistomatous leaf compared to an otherwise
identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
\aax{} be greater in sun than shade? We consider three nonmutually
exclusive hypotheses that we classify as `acclimation', `plasticity',
and `constitutive' (Fig.~\ref{fig-hypotheses}).

@@

\pandocbounded{\includegraphics[keepaspectratio]{../figures/hypotheses.pdf}}

@@

\caption{\label{fig-hypotheses}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimatopn hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdqty{2000}) than low measurement light
intensity (\ppfdequals{150}), regardless of growth light intensity or
native plant area index (PAI). The plasticity hypothesis predicts that
\aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}
